---
title: "Step intensities over arbitrary intervals"
author: "TA Trikalinos"
date: "2026-08-21"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Step intensities over arbitrary intervals}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
set.seed(20260821)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
K <- 10^5
```

## What this vignette shows

The vectorized samplers for piecewise constant intensity functions,
`vdraw_sc_step_regular()` and its zero-truncated counterpart, assume that the
rate matrix columns correspond to *equal-length* time intervals.
`vdraw_sc_step()` and `vztdraw_sc_step()` drop this assumption: the interval
bounds are passed in `time_breaks`, either once for all point processes or
separately for each one. Sampling remains exact -- the inversion algorithm does
not care whether the intervals are equally spaced.

This matters in practice because piecewise constant rates are rarely published
on equal-length intervals. The standard demographic and epidemiologic
convention uses age bands $[0, 1), [1, 5)$, five-year bands up to age $85$, and
a terminal open-ended band. With `vdraw_sc_step_regular()` one has to expand
such rates to a fine regular grid (feasible here, with integer bounds, but
wasteful; impossible exactly when the bounds are not commensurate). With
`vdraw_sc_step()` the published bands are used as-is.

We showcase the new functions and then do a small timing study against the
regular-grid sampler.

## Simulation description

Assume a population of $K = `r K`$ individuals indexed by $k \in [K] := \{1, \dots, K\}$.
Each is subject to an age-specific disease incidence
$\lambda_k(t) = w_k \, \lambda(t)$, where $\lambda(t)$ is a step function over
the $M = 19$ standard age bands and $w_k$ is a person-specific
multiplier (frailty) with $w_k \sim \textrm{Gamma}(5, 5)$, so that
$E[w_k] = 1$.

```{r setup}
library(nhppp)

age_breaks <- c(0, 1, seq(5, 85, by = 5), 100) # M + 1 = 20 bounds
lambda_bands <- c(
  0.5, 0.3, 0.2, 0.3, 0.5, 0.8, 1.2, 2, 3.5, 6,
  11, 20, 35, 60, 95, 140, 190, 235, 270
) / 10^5 # rates per person-year, one per band

w <- stats::rgamma(K, shape = 5, rate = 5)
lambda_matrix <- outer(w, lambda_bands) # K x M
```

The expected number of events over a full life course is
$w_k \int_0^{100} \lambda(t) \, dt = w_k \times `r round(sum(lambda_bands * diff(age_breaks)), 4)`$,
so most people have no event and few have more than one.

## Sampling with shared interval bounds

When all point processes share the same interval bounds, `time_breaks` is a
vector with `ncol(lambda_matrix) + 1` increasing values. Each row of the
returned matrix holds the (ordered) event times of one point process; rows
with fewer events than the widest row are padded with `NA`.

```{r whole-range}
Z <- vdraw_sc_step(
  lambda_matrix = lambda_matrix,
  time_breaks = age_breaks
)
dim(Z)
mean(!is.na(Z[, 1])) # proportion of people with at least one event
```

As with the other `vdraw_*()` functions, `Lambda_matrix` (cumulative
intensities at the interval ends) can be given instead of `lambda_matrix`,
and `atmost1` / `atmostB` / `atleast1` work the same way.

### The sampled times follow the target intensity

Pooled over people, the expected density of event ages is proportional to
$\sum_k w_k \lambda(t) \propto \lambda(t)$ -- the frailties scale the counts,
not the shape. The histogram of all simulated event ages should match
$\lambda(t) / \int_0^{100}\lambda(u)\,du$, a step function. It does; there is
no approximation at play.

```{r histogram, fig.alt="Histogram of simulated event ages overlaid with the normalized step intensity function. The two agree."}
if (nchar(system.file(package = "ggplot2")) > 0) {
  library(ggplot2)
  t_events <- as.vector(Z)
  t_events <- t_events[!is.na(t_events)]
  dens_bands <- lambda_bands / sum(lambda_bands * diff(age_breaks))
  step_df <- data.frame(
    age = age_breaks,
    density = c(dens_bands, dens_bands[length(dens_bands)])
  )
  ggplot() +
    geom_histogram(
      aes(x = t_events, y = after_stat(density)),
      breaks = 0:100, fill = "gray", color = "white"
    ) +
    geom_step(
      data = step_df, aes(x = age, y = density),
      direction = "hv", color = "red"
    ) +
    labs(
      x = "Age (years)",
      y = "Empirical and theoretical density",
      title = "All simulated event ages"
    )
}
```

## Per-process interval bounds

`time_breaks` can also be a matrix with one row of bounds per point process.
The rows of `lambda_matrix` still refer to the same bands -- but each band is
located at different times for different people.

A natural example: simulate in *calendar time* rather than age. If person $k$
is born in year $B_k$, their age bands correspond to the calendar intervals
$B_k + [0, 1), B_k + [1, 5), \dots$ We pass those directly.

```{r per-row-breaks}
birth_year <- stats::runif(K, min = 1920, max = 2000)
calendar_breaks <- outer(birth_year, age_breaks, FUN = "+") # K x (M + 1)

Z_cal <- vdraw_sc_step(
  lambda_matrix = lambda_matrix,
  time_breaks = calendar_breaks
)
range(Z_cal, na.rm = TRUE) # calendar years, not ages
```

Subtracting each person's birth year recovers event ages with the same
distribution as before (different draws, same process):

```{r qq-per-row, fig.alt="QQ plot comparing event ages sampled on the age scale and on the calendar-time scale. The agreement is excellent."}
age_at_event <- as.vector(Z_cal - birth_year) # birth_year recycles down columns
qqplot(
  as.vector(Z), age_at_event,
  xlab = "Event ages, shared breaks",
  ylab = "Event ages, per-person calendar breaks"
)
abline(0, 1, col = "red")
```

## Subintervals and first events

The rate information often spans a wider interval than the one to be sampled.
With `t_min` and `t_max` (scalars or vectors), events are drawn only in
`(t_min, t_max]`, without rebuilding the rate matrix. Here we draw each
person's *first* event between study entry at age 40 and a person-specific
exit age $T_{k1} \sim U(60, 100)$:

```{r subinterval}
T1 <- stats::runif(K, min = 60, max = 100)
t_first <- vdraw_sc_step(
  lambda_matrix = lambda_matrix,
  time_breaks = age_breaks,
  t_min = 40,
  t_max = T1,
  atmost1 = TRUE
)
summary(as.vector(t_first))
```

## Zero-truncated sampling

`vztdraw_sc_step()` -- equivalently `vdraw_sc_step(atleast1 = TRUE)` --
conditions on at least one event in the sampled interval. This is the
natural tool when simulating people who are known to become cases, e.g.,
drawing an age at diagnosis for each of $K$ cases:

```{r zero-truncated}
t_dx <- vztdraw_sc_step(
  lambda_matrix = lambda_matrix,
  time_breaks = age_breaks,
  t_min = 40,
  t_max = T1,
  atmost1 = TRUE
)
sum(is.na(t_dx)) # everyone has an event now
```

## Timing study

How much does the extra generality cost? On a regular grid the sampler maps a
time to its interval with one division; on arbitrary bounds it needs a binary
search over the breaks. And the band representation is more compact: $19$
columns instead of the $100$ one-year columns that `vdraw_sc_step_regular()`
would need to represent the same step function exactly. (A one-year grid works
here because the band bounds are integers. With non-commensurate bounds no
regular grid is exact.)

We compare four exact ways to sample the same process, on
$K_2 = 10^4$ point processes:

1. `vdraw_sc_step()`, the $19$ bands, shared `time_breaks` vector;
2. `vdraw_sc_step()`, the $19$ bands, per-row `time_breaks` matrix
   (identical rows, to isolate the cost of per-row bounds);
3. `vdraw_sc_step()`, the same step function expanded to a one-year grid
   ($100$ columns), to isolate the cost of the binary search at equal width;
4. `vdraw_sc_step_regular()` on the one-year grid -- the only exact option
   before `vdraw_sc_step()` existed.

First check that methods 3 and 4 are not just distributionally equal: on a
regular grid the general kernel consumes the random-number stream exactly like
the regular kernel, so the same seed gives the same draws (up to floating-point
roundoff in the internal construction of the cumulative intensities).

```{r same-seed}
K2 <- 10^4
lam_bands_2 <- lambda_matrix[1:K2, ]
lam_grid_2 <- lam_bands_2[, rep(seq_along(lambda_bands), times = diff(age_breaks))]
dim(lam_grid_2)

set.seed(2026)
Z_general <- vdraw_sc_step(
  lambda_matrix = lam_grid_2, time_breaks = 0:100
)
set.seed(2026)
Z_regular <- vdraw_sc_step_regular(
  lambda_matrix = lam_grid_2, rate_matrix_t_min = 0, rate_matrix_t_max = 100
)
all.equal(Z_general, Z_regular, tolerance = 10^-12)
```

Now the timings: median elapsed seconds over `R = 10` replications.
(The absolute numbers depend on the machine that rendered this document;
the relative comparisons are what matter.)

```{r timing}
median_elapsed <- function(f, R = 10) {
  stats::median(
    vapply(seq_len(R), function(i) system.time(f())[["elapsed"]], numeric(1))
  )
}

breaks_rowwise <- matrix(rep(age_breaks, each = K2), nrow = K2)

timing <- data.frame(
  method = c(
    "vdraw_sc_step, bands, shared breaks",
    "vdraw_sc_step, bands, per-row breaks",
    "vdraw_sc_step, one-year grid",
    "vdraw_sc_step_regular, one-year grid"
  ),
  columns = c(19, 19, 100, 100),
  seconds = c(
    median_elapsed(function() {
      vdraw_sc_step(lambda_matrix = lam_bands_2, time_breaks = age_breaks)
    }),
    median_elapsed(function() {
      vdraw_sc_step(lambda_matrix = lam_bands_2, time_breaks = breaks_rowwise)
    }),
    median_elapsed(function() {
      vdraw_sc_step(lambda_matrix = lam_grid_2, time_breaks = 0:100)
    }),
    median_elapsed(function() {
      vdraw_sc_step_regular(
        lambda_matrix = lam_grid_2,
        rate_matrix_t_min = 0, rate_matrix_t_max = 100
      )
    })
  )
)
knitr::kable(timing, digits = 4)
```

Three observations:

* The dominant cost is the width of the rate matrix: both samplers spend most
  of their time constructing cumulative intensities column by column. Sampling
  the $19$ bands directly is a few times faster than either sampler on the
  expanded $100$-column grid -- and the rate matrix is about a fifth of the
  size.
* The generality is not free. At equal width (methods 3 vs 4), the general
  kernel pays for the binary search over the breaks (instead of a division)
  and for reconstructing event times from the breaks; on our machines it takes
  roughly $1.5$ times as long as the regular kernel. When the grid really is
  regular, `vdraw_sc_step_regular()` remains the faster choice.
* Per-row `time_breaks` (method 2) cost little more than a shared vector.
  Note that a shared vector (or, equivalently, a 1-row matrix) is *not*
  replicated internally: the C++ kernel reads the single row of bounds for
  every point process, so there is no hidden $K \times (M+1)$ allocation.

## Bibliography

Trikalinos TA, Sereda Y. _nhppp: Simulating Nonhomogeneous Poisson Point Processes in R_. arXiv preprint arXiv:2402.00358. 2024 Feb 1.

Trikalinos TA, Sereda Y. _The nhppp package for simulating non-homogeneous Poisson point processes in R_. PLoS One. 2024 Nov 21;19(11):e0311311.

Since the publication of the paper, the syntax and options of the `nhppp` package have
evolved. To reproduce the code in the paper, you have to install the version
of `nhppp` used in the paper. Alternatively, take a look at the vignettes,
which are written to work with the current package.
