---
title: "SMDM2022_presentation"
output:
  pdf_document: default
  html_document: default
date: "2022-10-22"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(rstream)
library(ggplot2)
library(gganimate)
library(latex2exp)
library(des.sim.functions)
library(numDeriv)
library(nhppp)

n_times <- 1000
save_to <- file.path("~/Downloads/SMDM/2022")
```

## The functions for $\lambda(t)$, $\Lambda(t)$, and $\Lambda^{-1}(t)$

Use the `func_factory_cohort_annual_mort_rate()` function of the `des.sim.functions` 
package to create empirical functions for the $\lambda(t)$, $\Lambda(t)$, and $\Lambda^{-1}(t)$ of all cause mortality for a woman born in 1975 in the US.  

```{r }
# RI life tables (USMDB)
L_list <- func_factory_cohort_annual_mort_rate(
  sex = "male",
  cohort = 1975,
  race = "all",
  region = "USA",
  source = "HMD",
  interpolation_method = "linear"
)

# for the age_range in the returned functions, create a range of ages;
# to make the plot seem smoother, you specify a smaller value for `by`, e.g.
# `by = 0.1`.
age <- seq.int(
  from = L_list$age_range[1],
  to = L_list$age_range[2],
  by = 1
)

# Add data for plotting in the the tibble `df`
df <- as_tibble_col(age, column_name = "age") %>%
  mutate(
    lambda = L_list$lambda(age),
    Lambda = L_list$Lambda(age),
    Lambda_inv = L_list$Lambda_inv(age)
  )
```

### Intensity function $\lambda(t)$
```{r}
p1 <- ggplot(data = df, aes(x = age, y = lambda)) +
  geom_smooth(formula = "y ~ x", se = FALSE, method = "loess", span = 0.1) +
  labs(x = "Age", y = "Intensity") +
  theme_bw() +
  #  ggtitle("Intensity function") +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.title = element_text(size = 20),
    title = element_text(size = 18)
  )
# annotate("text", label = TeX("$lambda(t)$"), x=10, y=0.05, size=10)


show(p1)
```

### Cumulative intensity function $\Lambda(t)$
```{r}
p2 <- ggplot(data = df, aes(x = age, y = Lambda)) +
  geom_smooth(formula = "y ~ x", se = FALSE, method = "loess", span = 0.1) +
  labs(x = "Age", y = "Cumulative Intensity") +
  theme_bw() +
  # ggtitle("Cumulative intensity function") +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.title = element_text(size = 20),
    title = element_text(size = 18)
  ) #+
# annotate("text", label = TeX("$Lambda(t)$"), x=10, y=1.3, size=10)

show(p2)
```
### Inverse of cumulative intensity function $\Lambda^{-1}(t)$
```{r}
p2_inv <- ggplot(data = df, aes(x = age, y = Lambda_inv)) +
  geom_smooth(formula = "y ~ x", se = FALSE, method = "loess", span = 0.1) +
  labs(y = "Age", x = "Cumulative Intensity") +
  theme_bw() +
  # ggtitle("Cumulative intensity function") +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.title = element_text(size = 20),
    title = element_text(size = 18)
  ) +
  annotate("text", label = TeX("$Lambda^{-1}(t)$"), x = 15, y = 30, size = 10)

show(p2_inv)
```


# Computation time study

```{r}
# Parameters
S <- new("rstream.mrg32k3a")
l <- function(t) {
  2 * t
}
L <- function(t) {
  t^2
}
Li <- function(z) {
  sqrt(z)
}
tp <- c(0, 5)
t <- seq(tp[1], tp[2], length.out = 1000)
l_max <- max(sapply(t, l))

par(mfrow = c(1, 3))
plot(t, l(t))
plot(t, L(t))
plot(t, Li(t))
```


### `R` code: Best case arguments: RNG vs no RNG

```{r}
# Best case: RNG with no RNG
res1 <- microbenchmark::microbenchmark(
  # using the rstream random number generator
  "R: Inversion (substream)" = des.sim.functions::sim_nhppp_t_inv(
    Lambda = L, Lambda_inv = Li, range_t = tp, rng_stream = S
  ),
  "R: Thinning (substream)" = des.sim.functions::sim_nhppp_t_thinning(
    lambda = l, lambda_max = l_max, range_t = tp, rng_stream = S
  ),

  # using the built-in random number generator
  "R: Inversion (native)" = des.sim.functions::sim_nhppp_t_inv(
    Lambda = L, Lambda_inv = Li, range_t = tp, rng_stream = NULL
  ),
  "R: Thinning (native)" = des.sim.functions::sim_nhppp_t_thinning(
    lambda = l, lambda_max = l_max, range_t = tp, rng_stream = NULL
  ),
  times = n_times
)

autoplot(res1, type = c("violin")) +
  labs(x = "", y = "") +
  ggtitle("Best arguments, native RNG vs separate RNG substreams") +
  theme(
    axis.text.x = element_text(size = 14),
    axis.text.y = element_text(size = 14)
  )
```

### `R` code: Best case arguments: 1 sample vs all samples
```{r}

# Best case: 1 vs all
res2 <- microbenchmark::microbenchmark(
  # First time
  "R: Inversion (1 sample)" = des.sim.functions::sim_nhppp_t_inv(
    Lambda = L, Lambda_inv = Li, range_t = tp, rng_stream = NULL, only1 = T
  ),
  "R: Thinning (1 sample)" = des.sim.functions::sim_nhppp_t_thinning(
    lambda = l, lambda_max = l_max, range_t = tp, rng_stream = NULL, only1 = T
  ),

  # All times
  "R: Inversion (all samples)" = des.sim.functions::sim_nhppp_t_inv(
    Lambda = L, Lambda_inv = Li, range_t = tp, rng_stream = NULL, only1 = F
  ),
  "R: Thinning (all samples)" = des.sim.functions::sim_nhppp_t_thinning(
    lambda = l, lambda_max = l_max, range_t = tp, rng_stream = NULL, only1 = F
  ),
  times = n_times
)

autoplot(res2, type = c("violin")) +
  labs(x = "", y = "") +
  ggtitle("Best arguments, 1 sample vs all samples") +
  theme(
    axis.text.x = element_text(size = 14),
    axis.text.y = element_text(size = 14)
  )
```

### `R` code: Suboptimal case arguments: thinning vs inversion
```{r}
# Worse case:
inversion_w_integration <- function(lambda, range_t, only1 = FALSE) {
  Lambda_int <- function(t) {
    des.sim.functions::simpson_num_integr(f = lambda, a = range_t[1], b = t, n = 50)
  }
  des.sim.functions::sim_nhppp_t_inv(Lambda = Lambda_int, range_t = range_t, rng_stream = NULL, only1 = only1)
}

thinning_with_differentiation <- function(Lambda, range_t, only1 = FALSE) {
  lambda_diff <- function(t) {
    numDeriv::grad(Lambda, t)
  }
  des.sim.functions::sim_nhppp_t_thinning(
    lambda = lambda_diff, lambda_max = NULL, range_t = range_t, rng_stream = NULL, only1 = only1
  )
}

# The first time vs all
res3 <- microbenchmark::microbenchmark(
  "R: Inversion (1 sample)" = inversion_w_integration(lambda = l, range_t = tp, only1 = TRUE),
  "R: Thinning (1 sample)" = thinning_with_differentiation(Lambda = L, range_t = tp, only1 = TRUE),
  "R: Inversion (all samples)" = inversion_w_integration(lambda = l, range_t = tp, only1 = FALSE),
  "R: Thinning, (all samples)" = thinning_with_differentiation(Lambda = L, range_t = tp, only1 = FALSE),
  times = n_times
)

autoplot(res3, type = c("violin")) +
  labs(x = "", y = "") +
  ggtitle("Suboptimal arguments, 1 sample vs all samples") +
  theme(
    axis.text.x = element_text(size = 14),
    axis.text.y = element_text(size = 14)
  )
```

### Comparisons with `C++` code via `RCpp`

```{r}
res4 <- microbenchmark::microbenchmark(
  "C++: Inversion (one sample)" = nhppp::sim_nhppp_ct_inv(0, 5, "L", "Linv", TRUE),
  "C++: Thinning (one sample)" = nhppp::sim_nhppp_ct_thinning(0, 5, 10, "l", TRUE),
  "C++: Inversion (all samples)" = nhppp::sim_nhppp_ct_inv(0, 5, "L", "Linv", FALSE),
  "C++: Thinning (all samples)" = nhppp::sim_nhppp_ct_thinning(0, 5, 10, "l", FALSE),
  times = n_times
)

autoplot(res4, type = c("violin")) +
  labs(x = "", y = "") +
  ggtitle("C++ code, 1 sample vs all samples") +
  theme(
    axis.text.x = element_text(size = 14),
    axis.text.y = element_text(size = 14)
  )
```
### Single plot to choose from
```{r}
autoplot(rbind(res2, res4) %>% filter(str_detect(expr, "(all samples)")))
```


